Interpretable Casual Interventions \\ Encourage Core Risk Minimization

### 2025-02-03 paper/main.tex
switched hardinet to module-based invocation, updated paper title, drafted req. tables (w/o entries)
updated model weights, hardinet dataloader faulty `self.` call removed

\title{Contrastive Optimization -- Preliminaries}

@@

\begin{theorem}\label{abce}
	We can integrate backgroud invariance to the definition of binary cross-entropy using the following formulation:
	\begin{gather}
		\mathcal{J}(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
	\end{gather}
\end{theorem}
\begin{proof} For any logit $i$, binary cross-entropy loss is defined as:
	\begin{align}
		H(\theta, \mathcal{D}) &= -\left[ y_i \log (\sigma(\bm{s}_i)) + (1 - y_i) \log (1 - \sigma(\bm{s}_i)) \right] \\
		&= -\left[ y_i \log \left(\sigma\left(\sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) + (1 - y_i) \log \left(1 - \sigma\left(\sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) \right]
	\end{align}
	Similar to Theorem \ref{ace}, we can break down each HiResCAM to foreground and background components. For a non-target HiResCAM, anything that is not the target is inherently the background. Therefore, it's definition is retrained.
	\begin{gather}
		H(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
	\end{gather}
	The current formulation motivates activating either the foreground or the background for positive classification, and motivates de-activating every pixel of the non-positive class. We penalize activation on the background for the positive class only:
	\begin{gather}
		\mathcal{J}(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
	\end{gather}
	This gives us the ablated binary cross-entropy formulation.
\end{proof}

@@

@@

\paragraph{Reproducibility.} The source code, dataset, experiments, evals, and model weights can be found at \url{https://dagshub.com/jinensetpal/contrastive-optimization}.

@@

@@

@@

To test how well the approach generalizes out of the training distribution, we evaluate the performance of the binary classification contrastive model on random samples ($n=1000$) from the \href{https://www.microsoft.com/en-us/download/details.aspx?id=54765}{Kaggle Cats and Dogs Dataset}:
\begin{table}[h]
	\centering
	\begin{tabular}{c|c}
		\toprule
		\textbf{Method} & \textbf{Accuracy} \\
		\midrule
		Cross-Entropy & 77.0\% \\
		Ablated CE (Ours) & 75.7\% \\
		Ablated CE + KLD (Ours) & \bf 83.4\% \\
		\bottomrule
	\end{tabular}
\end{table}

@@

@@

\paragraph{Scaling the Approach} Training ResNet-50 on ImageNet and PASCAL VOC can allow us to evaluate if the approach scales effectively, with the performance gain not being constrained to the Oxford-IIIT Pets Dataset.

@@

\bibliographystyle{plain}
\bibliography{references}

>>>>

\title{Interpretable Casual Interventions \\
Encourage Core Risk Minimization}

@@

\begin{theorem}\label{abce}
	We can integrate backgroud invariance to the definition of binary cross-entropy using the following formulation:
	\begin{gather}
		\mathcal{J}(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
	\end{gather}
\end{theorem}
\begin{proof} For any logit $i$, binary cross-entropy loss is defined as:
	\begin{align}
		H(\theta, \mathcal{D}) &= -\left[ y_i \log (\sigma(\bm{s}_i)) + (1 - y_i) \log (1 - \sigma(\bm{s}_i)) \right] \\
		&= -\left[ y_i \log \left(\sigma\left(\sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) + (1 - y_i) \log \left(1 - \sigma\left(\sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) \right]
	\end{align}
	Similar to Theorem \ref{ace}, we can break down each HiResCAM to foreground and background components. For a non-target HiResCAM, anything that is not the target is inherently the background. Therefore, it's definition is retained.
	\begin{gather}
		H(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
	\end{gather}
	The current formulation motivates activating either the foreground or the background for positive classification, and motivates de-activating every pixel of the non-positive class. We penalize activation on the background for the positive class only:
	\begin{gather}
		\mathcal{J}(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
	\end{gather}
	This gives us the ablated binary cross-entropy formulation.
\end{proof}

@@

\subsubsection{Spatial Sliced Wasserstein Loss}

@@

\paragraph{Reproducibility.} The source code, dataset, experiments, evals, and model weights can be found at \url{https://dagshub.com/jinensetpal/contrastive-optimization.git}.

@@

\subsection{Overview}

We present training results on the following settings:
\begin{enumerate}
	\item \textbf{Oxford IIIT-Pets Dataset:} Binary Classification (Cat / Dog Classification) -- OOD Evaluation without semantic-shift. Multiclass Classification on 37-class fine-grained classification task.
	\item \textbf{Semantic Boundaries Dataset:} 20-class multilabel classification, evaluation on VOC2012 Test Set reporting AP, mIOU.
	\item \textbf{S-OOD ImageNet:} reporting in-distribution accuracy, 
\end{enumerate}

@@

\subsection{Benchmark: Hard-Imagenet}

\begin{table}[H]
	\centering
	\caption{Accuracies under various ablations, Relative Foreground Sensitivity, on FineTuned ResNet-50 models trained using varying approaches.}
	\begin{tabular}{c|c|ccc|c}
		\toprule
		\textbf{Method} & \textbf{None} $(\uparrow)$ & \textbf{Gray Mask FG} $(\downarrow)$ & \textbf{Gray BBOX FG} $(\downarrow)$ & \textbf{Tile FG} $(\downarrow)$ & \textbf{RFS} $(\uparrow)$ \\
		\midrule
		Cross-Entropy & 94.25 & 75.94 & 69.39 & 67.38 & -0.18 \\
		\midrule
		CORM & 92.91 & 76.20 & 69.12 & 68.32 & -0.08 \\
		DRF & \bf 94.39 & 73.53 & 67.51 & 66.71 & -0.27 \\
		CORM + DFR & 91.31 & 72.59 & 63.64 & 63.90 & -0.23 \\
		\midrule
		ACE (Ours) & 89.20 & 47.33 & 34.27 & 34.93 & 0.21 \\
		ACE + SWD (Ours) & 91.20 & \bf 38.27 & \bf 27.20 & \bf 33.60 & \bf 0.31 \\
		\bottomrule
	\end{tabular}
\end{table}

@@

\subsubsection{No Semantic Shift}
To test how well the approach generalizes out of the training distribution, we evaluate the performance of the binary classification contrastive model on random samples ($n=1000$) from the \href{https://www.microsoft.com/en-us/download/details.aspx?id=54765}{Kaggle Cats and Dogs Dataset}:
\begin{table}[h]
	\centering
	\begin{tabular}{c|ccc}
		\toprule
		\textbf{Method} & \textbf{Accuracy} & \textbf{Avg Conf | Misclassified} & \textbf{Avg Conf} \\
		\midrule
		Cross-Entropy 		& 77.0\% 	& TBR & TBR \\
		Ablated CE (Ours) 	& 75.7\% 	& TBR & TBR \\
		Ablated CE + KLD (Ours) & \bf 83.4\% 	& TBR & TBR \\
		Ablated CE + SWD (Ours) & TBR & TBR & TBR \\
		\bottomrule
	\end{tabular}
\end{table}

@@

\subsubsection{With Semantic Shift}

To evaluate the performance of the approach against semantic shifts, we can use the \citet{soodimagenet} dataset. It

The descriptors for abbreviations used in \ref{sood-wo-pretrain}, \ref{sood-w-pretrain} are as follows:
\begin{table}[h]
	\centering
	\begin{tabular}{c|c}
		\toprule
		\textbf{Key} & \textbf{Descriptor} \\
		\midrule
		F1 & F-1 Score \\
		ID & In-Distribution \\ 
		E & Easy Test Set -- Small Semantic Shift \\
		H & Hard Test Set -- Significant Semantic Shift \\
		AC & Average Confidence \\
		M & Misclassified Samples Only \\
		\bottomrule
	\end{tabular}
\end{table}

\begin{table}[h]
	\label{sood-wo-pretrain}
	\caption{Without pre-training:}
	\centering
	\hspace*{-2.5em}
	\begin{tabular}{c|ccccccc}
		\toprule
		\textbf{Method} & \textbf{F1 | ID} & \textbf{F1 | E} & \textbf{F1 | H} & \textbf{AC | M, E} & \textbf{AC | E} & \textbf{AC | M, H} & \textbf{AC | H} \\
		\midrule
		Cross-Entropy 		& TBR & 69.6\% & 52.9\% & .871 & .725 & .833 & .741 \\
		Ablated CE (Ours) 	& TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
		Ablated CE + KLD (Ours) & TBR & 72.0\% & 55.9\% & .909 & .790  & .879 & .805 \\
		Ablated CE + SWD (Ours) & TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
		\bottomrule
	\end{tabular}
\end{table}

\begin{table}[h]
	\label{sood-w-pretrain}
	\caption{With pre-training:}
	\centering
	\hspace*{-2.5em}
	\begin{tabular}{c|ccccccc}
		\toprule
		\textbf{Method} & \textbf{F1 | ID} & \textbf{F1 | E} & \textbf{F1 | H} & \textbf{AC | M, E} & \textbf{AC | E} & \textbf{AC | M, H} & \textbf{AC | H} \\
		\midrule
		Cross-Entropy 		& TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
		Ablated CE (Ours) 	& TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
		Ablated CE + KLD (Ours) & TBR & 78.1\% & 61.7\% & .904 & .747 & .864 & .766 \\
		Ablated CE + SWD (Ours) & TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
		\bottomrule
	\end{tabular}
\end{table}

\subsection{Qualitative Evals}

\subsubsection{Adversarial Robustness}

\subsubsection{Automated Image Circuit Discovery}

@@

@@

\bibliographystyle{plainnat}
\bibliography{references}
